A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-18 tex/body.tex
more

\begin{figure}[t!]
\normalsize
\centering
\begin{tabular}{c}
\(
\inferrule{
  t\typ{\delta} \in \tEc
  \betweenantes
  \Ctr^\delta = \{ \const{C}_1, \ldots, \const{C}_m \}
\\
  \bigl( \const s( t ) \in \tEc \text{ and } \const s \in \Sel^\delta \bigr)
  \text{ or }
  \bigl( \delta \in \Data \text{ and } \left|\delta\right| \text{ is finite} \bigr)
}{
  \Ec := \Ec,\: t \teq \const{C}_1\bigl(\const s^1_1( t ), \ldots,\const s^{n_1}_1( t ) \bigr) %\ROR
  \ROR \cdots \ROR \Ec := \Ec,\: t \teq \const{C}_m\bigl(\const s^1_m( t ), \ldots,\const s^{n_m}_m( t ) \bigr)
}
\)
\rn{Split}
\\[5\jot]
\(
\inferrule{
  t\typ{\delta}, u\typ{\delta} \in \tEc
  \betweenantes
  \delta \in \Codata
  \betweenantes
  \left|\delta\right| = 1
}{
  \Ec := \Ec,\: t \teq u
}
\)
\rn{Single}
\end{tabular}
%%% TODO: Clarify s's?
\caption{\,Derivation rules for branching%.
%All pairs of terms whose type has cardinality 1 are entailed to be equal (the degenerate case).
%Constructors must be assigned for all terms $t$ if has a selector is applied to it, or if $t$ has finite datatype type.
}
\label{fig:split-rule}
\end{figure}

@@

Assume $\Ec$ is initially the set $M_i$ described above. For each equality $t \teq s$ added
to $\Ec$, we associate a set of equalities from $M_i$ that together entail $t
\teq s$, which we call its \emph{explanation}.
Similarly, each $\Val \ec{x}$ is also associated an explanation---that is, a set
of equalities from $M_i$ that entail that the values of $\ec{x}$ in models of
$\Ec$ are of the form $\Val \ec{x}$. For example, if $x \teq \const{C}( x ) \in
M_i$, then $x \teq \const{C}( x )$ is an %(possible)
explanation for $\Val \ec{x} = \MU {\Varec{x}}.\; \const{C}( {\Varec{x}} )$.
Thus, the rules of the calculus are implemented as follows. For all rules with
conclusion $\bot$, we report the union of the explanations for all premises is
$\thD$-unsatisfiable. For \rn{Split}, we add a clause %of the form
$t \teq \const{C}_1\bigl( \const s^1_1( t ), \ldots, \const s^{n_1}_1( t ) \bigr)
\mathrel\lor \cdots \mathrel\lor t \teq \const{C}_m\bigl( \const s^1_m( t ),
\ldots, \const s^{n_m}_m( t ) \bigr)$ to $F$.
Decisions on which branch to take are thus performed externally by the SAT
solver. % and not by the theory solver.
All other rules add equalities to the
internal state of the theory solver. The rules in phase~1 are performed
eagerly---that is, for partial satisfying assignments $M$---while the rules in
phases~2 and~3 are performed only for complete satisfying
assignments $M$.

@@

However, the presence of quantified formulas poses an additional challenge in the context of (co)datatypes.
Namely, since quantified formulas may entail cardinality constraints on uninterpreted types $U$,
we may no longer assume that $U$ is of infinite cardinality.
When assuming that a finite type has infinite cardinality, the calculus presented in this paper is \emph{incomplete},
since the preconditions on \rn{Split} and \rn{Single} may not apply when they indeed are applicable.
This does not necessarily impact the correctness of our procedure in this setting,
since by assumption the solver is already incomplete when quantified formulas are present.
Regardless, we may be interested in increasing the precision of the solver by doing the following.
First, we may apply \rn{Split} to terms of datatype types whose cardinality depends on the finiteness of uninterpreted types.
Second, we may conditionally apply \rn{Single} to terms of codatatype types that may have cardinality one.
For instance, consider the codatatype $\keyw{codatatype}~\ty{c} = \const{C}(\ty{U},\: \ty{c})$, where $U$ is an uninterpreted type.
This codatatype has cardinality one when $\ty{U}$ has cardinality one.
If there exists two equivalence classes $\ec{t\typ{c}}$ and $\ec{u\typ{c}}$,
our implementation adds a clause of the form $( \exists x\,y.\, x \not\teq y \vee t \teq u )$ to $F$,
where $x$ and $y$ are variables of type $U$.
In other words, either the cardinality of $U$ is greater than one, or $t$ must be equal to $u$.

>>>>

\begin{figure}[t!]
\normalsize
\centering
\begin{tabular}{c}
\(
\inferrule{
  t\typ{\delta} \in \tEc
  \betweenantes
  \Ctr^\delta = \{ \const{C}_1, \ldots, \const{C}_m \}
\\
  \bigl( \const s( t ) \in \tEc \text{ and } \const s \in \Sel^\delta \bigr)
  \text{ or }
  \bigl( \delta \in \Data \text{ and } \left|\delta\right| \text{ is finite} \bigr)
}{
  \Ec := \Ec,\: t \teq \const{C}_1\bigl(\const s^1_1( t ), \ldots,\const s^{n_{1\!}}_1( t ) \bigr) %\ROR
  \ROR \cdots \ROR \Ec := \Ec,\: t \teq \const{C}_m\bigl(\const s^1_m( t ), \ldots,\const s^{n_m}_m( t ) \bigr)
}
\)
\rn{Split}
\\[5\jot]
\(
\inferrule{
  t\typ{\delta}, u\typ{\delta} \in \tEc
  \betweenantes
  \delta \in \Codata
  \betweenantes
  \left|\delta\right| = 1
}{
  \Ec := \Ec,\: t \teq u
}
\)
\rn{Single}
\end{tabular}
%%% TODO: Clarify s's?
\caption{\,Derivation rules for branching%.
%All pairs of terms whose type has cardinality 1 are entailed to be equal (the degenerate case).
%Constructors must be assigned for all terms $t$ if has a selector is applied to it, or if $t$ has finite datatype type.
}
\label{fig:split-rule}
\end{figure}

@@

Assume $\Ec$ is initially the set $M_i$ described above. For each equality $t \teq s$ added
to $\Ec$, we associate a set of equalities from $M_i$ that together entail $t
\teq s$, which we call its \emph{explanation}.
Similarly, each $\Val \ec{x}$ is also associated an explanation---that is, a set
of equalities from $M_i$ that entail that the values of $\ec{x}$ in models of
$\Ec$ are of the form $\Val \ec{x}$. For example, if $x \teq \const{C}( x ) \in
M_i$, then $x \teq \const{C}( x )$ is an %(possible)
explanation for $\Val \ec{x} = \MU {\Varec{x}}.\; \const{C}( {\Varec{x}} )$.
Thus, the rules of the calculus are implemented as follows. For all rules with
conclusion $\bot$, we report the union of the explanations for all premises is
$\thD$-unsatisfiable. For \rn{Split}, we add a clause %of the form
$t \teq \const{C}_1\bigl( \const s^1_1( t ), \ldots, \const s^{n_{1\!}}_1( t ) \bigr)
\mathrel\lor \cdots \mathrel\lor t \teq \const{C}_m\bigl( \const s^1_m( t ),
\ldots, \const s^{n_m}_m( t ) \bigr)$ to $F$.
Decisions on which branch to take are thus performed externally by the SAT
solver. % and not by the theory solver.
All other rules add equalities to the
internal state of the theory solver. The rules in phase~1 are performed
eagerly---that is, for partial satisfying assignments $M$---while the rules in
phases~2 and~3 are performed only for complete satisfying
assignments $M$.

@@

However, the presence of quantified formulas poses an additional challenge in the context of (co)datatypes.
Namely, since quantified formulas may entail cardinality constraints on uninterpreted types $\ty{u}$,
we may no longer assume that $\ty{u}$ is of infinite cardinality.
When assuming that a finite type has infinite cardinality, the calculus presented in this paper is \emph{incomplete},
since the preconditions on \rn{Split} and \rn{Single} may not apply when they indeed are applicable.
This does not necessarily impact the correctness of our procedure in this setting,
since by assumption the solver is already incomplete when quantified formulas are present.
Regardless, we may be interested in increasing the precision of the solver by doing the following.
First, we may apply \rn{Split} to terms of datatype types whose cardinality depends on the finiteness of uninterpreted types.
Second, we may conditionally apply \rn{Single} to terms of codatatype types that may have cardinality one.
For example, consider the codatatype $\keyw{codatatype}~\ty{c} = \const{C}(\ty{u},\: \ty{c})$, where $\ty{u}$ is an uninterpreted type.
This codatatype has cardinality one when $\ty{u}$ has cardinality one.
If there exist two equivalence classes $\ec{s\typ{c}}$ and $\ec{t\typ{c}}$,
our implementation adds the clause %of the form
$\exists x\;y\typ{\ty{u}}.\;\, x \not\teq y \mathrel\vee s \teq t$ to $F$.
In other words, either the cardinality of $\ty{u}$ is greater than one or $s$ is equal to $t$.
